Differential abundance analysis of taxonomically-biased microbiome measurements

### 2022-02-07 conclusion.Rmd
Revise conclusion
Improve previous text and add text on future directions.

_This section is a placeholder for a short (2-3 paragraph) conclusion. We are planning to keep 'discussion' to the earlier sections and appendix, at least for initial submission._

>>>>

It is commonly thought that analyses of the differences between samples are robust to the taxonomic bias inherent in MGS measurement as long as all samples have been subjected to the same MGS workflow.
But our results show that consistent taxonomic bias is still capable of creating scientific errors in common forms of DA analysis.
In particular, we showed mathematically that bias affects DA analysis methods based on proportions or in absolute abundances derived from them (using total-abundance normalization), due to variation in the mean efficiency across samples of varying taxonomic composition.
<!-- This is assuming consistent bias at the taxonomic level of analysis; things could be worse if we're analyzing higher taxonomic units than at which bias is conserved. -->
<!-- Perhaps state that we showed this theoretically, in mock communities, and in real case studies -->
The error, however, varies by experimental context and may often be negligible in a practical sense.
<!-- It is negligible if the mean efficiency is stable across samples, and may be unimportant if it is not correlated with the focal covariate in a regression analysis. -->
Moreover, it can be quantified and corrected through the use of reference-taxon measurements and/or community calibration controls.
In addition, bias-aware sensitivity and meta-analyses make it possible to rigorously account for bias even in the absence of such controls.
Applications of these methods will gradually improve our understanding of the experimental contexts in which variation in the mean efficiency is likely to be problematic and should be accounted for.

We also showed that other DA methods are more robust---perhaps even entirely invariant---to consistent taxonomic bias.
In particular, analyses based on multiplicative variation in ratios and of absolute abundances derived from them (using reference-species normalization) are invariant to bias in the MGS measurement, as bias causes a constant multiplicative error that cancels in cross-sample comparisons.
In addition, careful pairing of total-abundance and MGS measurement methods may remove most of the bias-driven error in absolute abundances from total-abundance normalization.

Important open questions and future research directions remain.
<!-- Taxonomic aggregation and non-FC analyses are left out because I'm assuming they will be addressed in the manuscript -->
Key open questions include 1) determining the extent to which bias is consistent across samples at the species level for a given MGS method and sample type; 2) assessing the validity of post-extraction absolute-abundance controls as measures of pre-extraction abundance; 3) understanding how the multiplicative error from bias interacts with the non-multiplicative error from contamination and taxonomic misassignment; and 4) understanding how our assumptions about underlying community dynamics, and in particular what information zero counts provide, affect bias sensitivity.
In addition, while we have showed some simple methods for incorporating bias and control measurements into DA analyses, better statistical tools are needed that are capable of including taxonomic bias and uncertainty in the MGS and supplemental non-MGS measurements.
Finally, more concrete experimental recommendations and user-friendly statistical workflows are needed for implementing the solutions we propose in various experimental contexts.

Our theoretical framework and example analyses provide a foundation for addressing these open questions and developing a practical experimental protocols and statistical tools.
We look to a future in which microbiome researchers routinely choose an appropriate combination of experimental and data-analytic methods capable of answering their fundamental question while accounting for taxonomic bias and the other inherent limitations of MGS measurements---and in this way, gaining the confidence needed to codify statistical findings into true scientific knowledge.
